DANA Normalization Assessment for Data from TCGA-BRCA and TCGA-UCS

### 2021-12-17 TCGA_BRCA_UCS_Data_Analysis.Rmd
Major Revision
Considerably improved the DANA method by replacing the metric for removal of handling effects by the mean-squared correlation reduction in negative controls. Updated figures and results

---
title: "Assessment of Normalization Methods for TCGA-BRCA and TCGA-UCS Data"
author: "Yannick Dueren, Johannes Lederer, Li-Xuan Qin"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
# abstract: "TODO: This is the abstract."
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "results") })
---

@@

The settings list `settings` configure which parts, simulations, and analysis steps of this R markdown are executed. This way single components of the analysis can be disabled for the purpose of saving computation time. Setting `settings$all` to `TRUE` will set all other options to true regardless of their value.

```{r analysis.settings}
settings <- list(
  # If true sets all other settings to TRUE
  all=F,
  # Generate interactive plots of co-expression structures
  investigate.coexpression=T,
  # Perform Differential Expression Analysis
  perform.DEA=T,
  # Generate paper figures
  generate.paper.figs=T,
  # Specify if paper figures are exported
  export.figures = F,
  # Path for file exports
  paper.fig.path = "../danawriteup/figs/",
  # Clears environment
  debug=TRUE
)
if(settings$all) {
  settings$investigate.coexpression=TRUE
  settings$perform.DEA=TRUE
  settings$generate.paper.figs=TRUE
}
if(settings$debug) rm(list=ls()[ls()!="settings"])
```

@@

# Apply DANA pipeline
DANA.results <- DANA(data.RC=data.RC, 
                     data.norm=data.norm, 
                     pos.controls=pos.controls, 
                     neg.controls=neg.controls, 
                     clusters=clusters.data,
                     coords=coords,
                     case.name="benchmark",
                     generate.plots=config$generate.plots,
                     preprocess.transform=config$preprocess.transform,
                     corr.method.pos=config$corr.method.pos,
                     tuning.criterion.pos=config$tuning.criterion.pos,
                     corr.method.neg=config$corr.method.neg,
                     tuning.criterion.neg=config$tuning.criterion.neg)

@@

  p.corr.pos.two <- arrangeGrob(p.corr.pos.DESeq + theme(legend.position="none"),
                                p.corr.pos + theme(legend.position="none"),
                                get.legend(p.corr.pos.DESeq + theme(legend.position = "bottom")),
                                layout_matrix=rbind(c(1,2), c(4,4)),
                                heights = c(5,1))
  grid.arrange(p.corr.pos.two)  # plot

@@

  p.corr.neg.two <- arrangeGrob(p.corr.neg.DESeq + theme(legend.position="none"),
                                p.corr.neg + theme(legend.position="none"),
                                get.legend(p.corr.neg.DESeq + theme(legend.position = "bottom")),
                                layout_matrix=rbind(c(1,2), c(4,4)),
                                heights = c(5,1))
  grid.arrange(p.corr.neg.two)  # plot

@@

### Subset for paper

```{r generate.paper.figs.corr.density}
if(settings$generate.paper.figs) {
  # Co-expression models
  model <- DANA.results$data.model
  norm.model1 <- DANA.results$norm.models$TCGA.DESeq
  norm.model2 <- DANA.results$norm.models$TCGA.RUVs

  # Set number of genes for negative correlation to minimum
  num.genes.plot.neg <- min(dim(model$neg$corr)[1],
                            dim(norm.model1$neg$corr)[1],
                            dim(norm.model2$neg$corr)[1])

  # Subsetted correlation matrices
  corr.neg <- model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.norm1.neg <- norm.model1$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.norm2.neg <- norm.model2$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]

  ## extract non-zero correlations
  nnz.corr.neg <- corr.neg[upper.tri(corr.neg)]
  nnz.corr.neg <- nnz.corr.neg[nnz.corr.neg != 0]
  nnz.corr.norm1.neg <- corr.norm1.neg[upper.tri(corr.norm1.neg)]
  nnz.corr.norm1.neg <- nnz.corr.norm1.neg[nnz.corr.norm1.neg != 0]
  nnz.corr.norm2.neg <- corr.norm2.neg[upper.tri(corr.norm2.neg)]
  nnz.corr.norm2.neg <- nnz.corr.norm2.neg[nnz.corr.norm2.neg != 0]

  # Keep positive correlations
  nnz.corr.neg <- nnz.corr.neg[nnz.corr.neg > 0]
  nnz.corr.norm1.neg <- nnz.corr.norm1.neg[nnz.corr.norm1.neg > 0]
  nnz.corr.norm2.neg <- nnz.corr.norm2.neg[nnz.corr.norm2.neg > 0]

  ## direct comparison of negative controls un-normalized vs. DESeq vs. RUVs
  neg.corr <- data.frame(
    control=factor(c(rep("Un-normalized", length(nnz.corr.neg)),
                     rep("DESeq", length(nnz.corr.norm1.neg)),
                     rep("RUVs", length(nnz.corr.norm2.neg)))),
    corr=abs(c(nnz.corr.neg, nnz.corr.norm1.neg, nnz.corr.norm2.neg))
    )
  # ttt <- data.frame(prop.table(table(method=neg.corr$control, corr=neg.corr$corr), 1))

  p.density.neg.nnz.freq <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.8, size=.9) +
    theme_classic() +
    xlim(0, 1) +
    # scale_linetype_manual(values=c("twodash", "solid", "solid", "solid"))+
    scale_color_manual(values=c('#e7298a','#1B9E77','#D95F02')) + # First colors from rcolorbrewer set "Dark2"
    theme(legend.position=c(0.9,0.86), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation")#+ ggtitle("Positive correlation among negative controls")
  print(p.density.neg.nnz.freq)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "BRCA_UCS_CorrDensityNeg_Freq.pdf"), p.density.neg.nnz.freq, width = 5, height=4, device="pdf")
  }
}
```

### All methods

```{r generate.paper.figs.corr.density.supplements}
if(settings$generate.paper.figs) {
  # Co-expression models
  model <- DANA.results$data.model

  # Set number of genes for negative correlation to minimum
  num.genes.plot.neg <- min(dim(model$neg$corr)[1],
                            sapply(DANA.results$norm.models, function(x) dim(x$neg$corr)[1]))

  # Subsetted correlation matrices
  corr.neg <- model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.norm.neg <- lapply(DANA.results$norm.models, function(x) x$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg])

  ## Strictly positive correlations
  nnz.corr.neg <- corr.neg[upper.tri(corr.neg)]
  nnz.corr.neg <- nnz.corr.neg[nnz.corr.neg > 0]
  nnz.corr.norm.neg <- lapply(corr.norm.neg, function(x) x[upper.tri(x)][x[upper.tri(x)] > 0])
  names(nnz.corr.norm.neg) <- substr(names(nnz.corr.norm.neg), 6, 20)

  control <- c(rep("un-normalized", length(nnz.corr.neg)))
  for(i in 1:length(nnz.corr.norm.neg)) {
    control <- c(control, rep(names(nnz.corr.norm.neg)[i], length(nnz.corr.norm.neg[[i]])))
  }
  ## direct comparison of negative controls
  neg.corr <- data.frame(
    control=factor(control),
    corr=abs(c(nnz.corr.neg, unname(unlist(nnz.corr.norm.neg))))
    )

  p.corr.frequency.neg.controls.all <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.9) +
    theme_classic() +
    xlim(0, 1) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) + # First 2 colors from rcolorbrewer set "Dark2"
    theme(legend.position=c(0.85,0.6), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") + # ggtitle("Positive correlation among negative controls") + 
    theme(legend.key.width = unit(2,"cm"))
  print(p.corr.frequency.neg.controls.all)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "BRCA_UCS_CorrDensityNeg_Freq_All.pdf"), p.corr.frequency.neg.controls.all, width = 8, height=5, device="pdf")
  }
}
```

@@

### Subset for paper

```{r generate.paper.figs.corr.scatter}
if(settings$generate.paper.figs) {
  ## Generate correlation scatter plots for positive controls for TMM and UQ
  par(mfrow=c(1,1))

  # un-normalized vs RUVr
  p.scatter.RUVr <- plot.corr.scatter(
    corr1=DANA.results$data.model$pos$corr, 
    corr2=DANA.results$norm.models$TCGA.RUVr$pos$corr,
    clusters=clusters.data,
    title="RUVr Normalization",
    xlab="un-normalized", 
    ylab="RUVr", 
    threshold=0)
  print(p.scatter.RUVr)
  # un-normalized vs Med
  p.scatter.Med <- plot.corr.scatter(
    corr1=DANA.results$data.model$pos$corr, 
    corr2=DANA.results$norm.models$TCGA.Med$pos$corr,
    clusters=clusters.data,
    title="Med Normalization",
    xlab="un-normalized", 
    ylab="Med", 
    threshold=0)
  print(p.scatter.Med)

  # Save plots
  if(settings$export.figures) {
    # ggsave(paste0(settings$paper.fig.path, "BRCA_UCS_ScatterRUVr.pdf"), p.scatter.RUVr, width=4, height=4, device="pdf")
    # ggsave(paste0(settings$paper.fig.path, "BRCA_UCS_ScatterMed.pdf"), p.scatter.Med, width=4, height=4, device="pdf")
  }
}
```

### All normalization methods for supplementary materials

@@

  test.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.results$metrics), 6, 20),
    cc=DANA.results$metrics[, "cc"],
    mcr=DANA.results$metrics[, "mcr"]
  )

  p.DANA.metrics <- ggplot(test.DANA.metrics,aes(x=mcr,y=cc, label=method)) + 
    geom_point(alpha=.5) +
    theme_classic() + 
    xlab(TeX("$mcr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3, max.overlaps = Inf, min.segment.length = 0.2) +
    # xlim(c(min(c(0, test.mposc.reduction)),max(test.mposc.reduction))) +
    scale_x_continuous(labels = scales::percent, limits=c(0,1.05), 
                       breaks = scales::pretty_breaks(n = 5)) +
    # ylim(c(0,1)) 
    scale_y_continuous(labels = scales::percent, limits = c(0,1))
  print(p.DANA.metrics)

  print("DANA Statistics for the TCGA-BRCA/UCS")
  test.DANA.metrics$cc <- round(test.DANA.metrics$cc,3)
  test.DANA.metrics$mcr <- round(test.DANA.metrics$mcr,3)
  print(test.DANA.metrics)

>>>>

---
title: "DANA Normalization Assessment for Data from TCGA-BRCA and TCGA-UCS"
author: "Yannick Dueren, Johannes Lederer, Li-Xuan Qin"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "results") })
---

@@

The settings list `settings` configure which parts, simulations, and analysis steps of this R markdown are executed. This way single components of the analysis can be disabled for the purpose of saving computation time.

```{r analysis.settings}
settings <- list(
  # Generate interactive plots of co-expression structures
  investigate.coexpression=F,
  # Perform Differential Expression Analysis
  perform.DEA=T,
  # Generate paper figures
  generate.paper.figs=T,
  # Specify if paper figures are exported
  export.figures = T,
  # Path for file exports
  paper.fig.path = "../danawriteup/figs/",
  # Clears environment
  debug=TRUE
)
if(settings$debug) rm(list=ls()[ls()!="settings"])
```

@@

# Apply DANA pipeline
DANA.results <- DANA(data.RC=data.RC, 
                     data.norm=data.norm, 
                     pos.controls=pos.controls, 
                     neg.controls=neg.controls, 
                     clusters=clusters.data,
                     coords=coords,
                     case.name="TCGA",
                     generate.plots=config$generate.plots,
                     preprocess.transform=config$preprocess.transform,
                     corr.method.pos=config$corr.method.pos,
                     tuning.criterion.pos=config$tuning.criterion.pos,
                     corr.method.neg=config$corr.method.neg,
                     tuning.criterion.neg=config$tuning.criterion.neg)

@@

  p.corr.pos.two <- arrangeGrob(p.corr.pos.DESeq + theme(legend.position="none"),
                                p.corr.pos + theme(legend.position="none"),
                                get.legend(p.corr.pos.DESeq + theme(legend.position = "bottom")),
                                layout_matrix=rbind(c(1,2), c(4,4)),
                                heights = c(5,1))
  grid.arrange(p.corr.pos.two)  # display plot

@@

  p.corr.neg.two <- arrangeGrob(p.corr.neg.DESeq + theme(legend.position="none"),
                                p.corr.neg + theme(legend.position="none"),
                                get.legend(p.corr.neg.DESeq + theme(legend.position = "bottom")),
                                layout_matrix=rbind(c(1,2), c(4,4)),
                                heights = c(5,1))
  grid.arrange(p.corr.neg.two)  # display plot

@@

```{r generate.paper.figs.corr.density.supplements}
if(settings$generate.paper.figs) {
  models <- list(
    "no-normalization" = DANA.results$data.model,
    TMM = DANA.results$norm.models$TCGA.TMM,
    DESeq = DANA.results$norm.models$TCGA.DESeq,
    PoissonSeq = DANA.results$norm.models$TCGA.PoissonSeq,
    RUVg = DANA.results$norm.models$TCGA.RUVg,
    RUVs = DANA.results$norm.models$TCGA.RUVs,
    RUVr = DANA.results$norm.models$TCGA.RUVr,
    TC = DANA.results$norm.models$TCGA.TC,
    Med = DANA.results$norm.models$TCGA.Med,
    UQ = DANA.results$norm.models$TCGA.UQ,
    QN = DANA.results$norm.models$TCGA.QN
  )

  # Set number of genes for negative correlation to minimum found
  num.genes.plot <- min(sapply(models, function(x) dim(x$neg$corr)[1]))

  # Subsetted correlation matrices
  corrs <- lapply(models, function(x) x$neg$corr[1:num.genes.plot, 1:num.genes.plot])
  corrs <- lapply(corrs, function(x) x[upper.tri(x)])

  control <- factor(
    as.vector(sapply(names(corrs), function(x) rep(x,length(corrs[[1]])))),
    levels = names(models) 
  )
  neg.corr <- data.frame(
    control=factor(control),
    corr=unname(unlist(corrs))
  )

  # plot insert
  p.insert <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_vline(xintercept = 0, size=.4, color="gray80", linetype="solid") +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.3) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) +
    theme_bw() +
    xlim(-1, 1) +
    theme(legend.position="none",
          axis.title.x = element_blank(),
          axis.title.y = element_blank())

  p.corr.frequency.neg.controls.all <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_vline(xintercept = 0, size=.4, color="gray80", linetype="solid") +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.75) +
    theme_classic() +
    xlim(-1, 1) +
    ylim(0,600) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) +
    theme(legend.position=c(0.85,0.8), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") + 
    theme(legend.key.width = unit(2.4,"cm")) +
    annotation_custom(ggplotGrob(p.insert), xmin=-1.1, xmax=-0.3, ymin=300, ymax=600)
  print(p.corr.frequency.neg.controls.all)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "BRCA_UCS_CorrDensityNeg_Freq_All.pdf"), p.corr.frequency.neg.controls.all, width = 8, height=8, device="pdf")
  }
}
```

@@

@@

  test.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.results$metrics), 6, 20),
    cc=DANA.results$metrics[, "cc"],
    mscr=DANA.results$metrics[, "mscr"]
  )

  p.DANA.metrics <- ggplot(test.DANA.metrics,aes(x=mscr,y=cc, label=method)) + 
    geom_point(alpha=.5) +
    theme_classic() + 
    xlab(TeX("$mscr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3, max.overlaps = Inf, min.segment.length = 0.3) +
    # xlim(c(min(c(0, test.mposc.reduction)),max(test.mposc.reduction))) +
    scale_x_continuous(labels = scales::percent, limits=c(0,1.05), 
                       breaks = scales::pretty_breaks(n = 5)) +
    # ylim(c(0,1)) 
    scale_y_continuous(labels = scales::percent, limits = c(0.5,1))
  print(p.DANA.metrics)

  print("DANA Statistics for the TCGA-BRCA/UCS")
  test.DANA.metrics$cc <- round(test.DANA.metrics$cc,3)
  test.DANA.metrics$mscr <- round(test.DANA.metrics$mscr,3)
  print(test.DANA.metrics)
